Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference Imaging Photometric Survey.}

### 2021-04-20 cluster_paper/ms.tex
update ms
Co-authored-by: Joel Hartman <jhartman@astro.princeton.edu>

\title{
  Cluster Difference Imaging Photometric Survey. III.
  Rotation and Lithium in an Open Cluster Spanning 500\,Parsecs
}

@@

To determine the reddening correction across \cn, we queried the 2MASS
DUST service\footnote{
  \url{http://irsa.ipac.caltech.edu/applications/DUST}; query
  performed using the \texttt{astrobase.services.dust} module
  \citep{bhatti_astrobase_2018}.  } and retrieved the extinction
parameters at the positions of each NGC\,2516 member.  The mean and
standard deviation of the $E(\mathrm{B}-\mathrm{V})$ values for the
\citet{schlegel_maps_1998} and \citet{schlafly_measuring_2011} maps
were consistent, so we adopted the average from
\citet{schlegel_maps_1998}: $E(\mathrm{B}-\mathrm{V})_{\rm
S}=0.206\pm0.039$.  \citet{bonifacio_search_2000} found however that
the \citet{schlegel_maps_1998} maps overestimate the reddening values
when the color excess exceeds about 0.10\,mag. Applying the correction
proposed by \citet{bonifacio_search_2000} leads to our adopted
$E(\mathrm{B}-\mathrm{V})=0.102\pm0.019$.  Finally, to convert to Gaia
colors we used the calibration of \citet{stassun_TIC8_2019}, namely
$E(\mathrm{Bp}-\mathrm{Rp})=1.31 E(\mathrm{B}-\mathrm{V})$ and
$A_G=2.72 E(\mathrm{B}-\mathrm{V})$.  This yielded
$E(\mathrm{Bp}-\mathrm{Rp})=0.134\pm0.025$, which is used in the
plots.  To redden the isochrones, we assumed $R_V=3.1$, and applied
the \citet{odonnell_1994} SED-dependent extinction law star-by-star
through the PARSEC interface.

For the metallicity, we considered a range of super and sub-solar
metallicities to fit as much of the locus from
$0.5<\mathrm{Bp}-\mathrm{Rp}<1.5$ as possible, and settled on the
slightly super-solar $[M/H]=0.06$ \citep{cummings_2011_li_iron}.
Sub-solar metallicities led to model predictions too blue along the
main sequence by $\approx$0.1\,mag.  Our adopted metallicity agrees
with the spectroscopic metallicity from
\citet[][Sec~4.4.4]{cummings_2011_li_iron}, and with the iron
abundance recently determined by the Gaia-ESO team
\citep{baratella_gaiaeso_2020}, which represented an up-revision from
an earlier sub-solar metallicity \citep{randich_gaiaeso_2018}.  While
\citet{bailey_rv_2018} report a slightly sub-solar metallicity for the
cluster, we prefer super-solar based on the goodness of fit from the
photometry.

@@

        \vspace{-0.5cm}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f3c.pdf}
			\includegraphics[width=0.49\textwidth]{f3d.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf TESS rotations periods and dereddened Gaia colors for
  \cn.} {\it Top row}: \cn\ split into its core and halo.  Sets
  $\mathcal{A}$ ({\it left}) and $\mathcal{B}$ ({\it right}) undergo
  successive stages of automated cleaning, (see
  Section~\ref{subsubsec:cluster}).  The Pleiades
  \citep[125\,Myr;][]{rebull_rotation_2016a} and Praesepe
  \citep[650\,Myr;][]{douglas_poking_2017} are shown for reference.
  Stars in the core and the halo of NGC\,2516 overlap with the
  Pleiades at $\bpmrp$$<$1.2.  From $1.2<\bpmrp<1.7$, some NGC\,2516
  members have longer rotation periods than Pleaides members.  {\it
  Bottom row}: \cn\ colored by binarity indicators.  Sources with
  $\mathrm{RUWE}>1.2$ appear in red (astrometric binaries; 11\% of
  stars).  Overplotted in orange are sources at least 0.3\,mag above
  an empirically fit cluster isochrone (photometric binaries; 20\% of
  stars).
  \label{fig:rot}
	}
\end{figure*}

@@

To clean the resulting rotation period measurements, we designed a set
of automated cleaning criteria, which yielded sets we denote
$\mathcal{A}$ and $\mathcal{B}$.  For both sets, we considered light
curves for which the peak Lomb-Scargle periodogram period was below 15
days, the normalized Lomb-Scargle power exceeded 0.08, and for which
no equal-brightness or greater companions were within the aperture.
Beyond requiring that the target star be the brightest star in the
aperture, we also required that at most one companion with brightness
exceeding one tenth of the target star's brightness be present in the
aperture according to the Gaia DR2 source catalog.  For set
$\mathcal{A}$, this yielded \nautorotnumerator\ light curves, out of
\nautorotdenominator\ stars for which rotation periods might plausibly
have been detected.  For set $\mathcal{B}$, we additionally required
that the Lomb-Scargle and Stellingwerf phase-dispersion methods
yielded identical periods to within 10\%.  This yielded
\nautorotnumeratormatching\ light curves.

@@

The results are shown in the top two rows of
Figure~\ref{fig:physical_x_rotn}.  In these figures, to ensure a fair
comparison, we only show stars with $0.5<\bpmrp<1.2$ for which our
TESS pipeline succeeding in making light curves.  In other words, the
stars in the base sample are those for which we could have plausibly
detected a rotation period.  This color range is preferred because the
the slow sequence becomes less defined for spectral types later than
$\approx$K4V.  The bins in the histograms (top-right) were chosen to
ensure that the same number of stars were in each bin.  Rotation
periods are detected for 66\% of stars at least 25\,pc from the
cluster core.  The average detection rate inside of 25\,pc, 73\%, is
slightly higher.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.9\textwidth]{f5a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown versus
  dereddened color for all kinematic \cn\ members with Gaia-ESO
  ($N=452$) or GALAH ($N=107$) spectra available.  The GALAH spectra
  comprise slightly over half of the halo stars, due to the
  non-targeted selection function of that survey.  Field stars (gray
  points) are Keck-HIRES measurements of stars with detected
  transiting planets from the Kepler mission
  \citep{berger_identifying_2018}.  Stars in the Pleiades (125\,Myr)
  with Li detections are from \citet{bouvier_pleiades_lirot_2018};
  Praesepe detections (650\,Myr) are shown in violet
  \citep{soderblom_praesepe_1993}.  Points with ${\rm
  EW}\approx0\,$m\AA\ are non-detections.
  {\bf LGB TODO: the increasing EWs for the Gaia-ESO M dwarfs must be erroneous. What's happening there?}
  \label{fig:lithiumcorehalo}
  }
\end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f5b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium and rotation in NGC\,2516.} Points show all
  kinematic \cn\ members with TESS rotation periods in the Gaia-ESO or
  GALAH surveys, and are colored by equivalent width.  Points with
  reported equivalent widths below 20\,m\AA\ (including
  non-detections) are shown with gray crosses.  The lithium-rotation
  correlation for the K-dwarfs has been observed in other young
  clusters, and is discussed in Section~\ref{subsec:lithium}.
	\label{fig:lithiumrot}
	}
\end{figure*}

@@

From their rotation period analysis of \cn,
\citet{fritzewski_rotation_2020} have recently demonstrated that
between 100 and 200\,Myr, young clusters (\cn, Blanco\,1, Psc\,Eri,
M\,35, M\,50, and the Pleaides) all show similar rotation period
distributions.  The FGK stars show a ``slow sequence'', a population
of ``rapid rotators'', and in some cases stars in the ``void'' between
the two populations ({\it e.g.,}, Psc\,Eri and \cn\ show more stars in
the void than the Pleiades).  The late M dwarfs with detected rotation
periods tend to rotate very rapidly ($<3$\,days), although these are
beyond the magnitude limit of our analysis.  This section briefly
discusses the origins of these features in the rotation-color diagram.

@@

The other features of the rotation-color diagram are indications that
magnetic braking is not always the most important effect.  The
question of why the rapid rotators exist, for instance, has not yet
been definitively answered.  Models that introduce variable
core-envelope decoupling timescales can reproduce each type of
population \citep{Irwin_NGC2516_2007}, but the question then shifts to
why these decoupling timescales should be differ for stars that are
otherwise identical.  Another possible explanation invokes a
spontaneous and random change of magnetic field topology that causes
the dynamo to couple more strongly with the stellar wind
\citep{brown_metastable_2014}.  Other processes could be external to the
star entirely: \citet{qureshi_signature_2018} for instance have
reported that mergers between giant planets and the star could explain
a non-neglible fraction of the rapid rotators.

>>>>

% Cluster Difference Imaging Photometric Survey. III.
% Rotation and Lithium in an Open Cluster Spanning 500\,Parsecs
\title{
  Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open
  Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference
  Imaging Photometric Survey.}
}

@@

To determine the reddening correction across \cn, we queried the 2MASS
DUST service\footnote{
  \url{http://irsa.ipac.caltech.edu/applications/DUST}; query
  performed using the \texttt{astrobase.services.dust} module
  \citep{bhatti_astrobase_2018}.  } and retrieved the total
line-of-sight extinction parameters at the positions of each \cn\ 
member.  The mean and standard deviation of the
$E(\mathrm{B}-\mathrm{V})$ values for the \citet{schlegel_maps_1998}
and \citet{schlafly_measuring_2011} maps were consistent, so we
adopted the average from \citet{schlegel_maps_1998}:
$E(\mathrm{B}-\mathrm{V})_{\rm S}=0.206\pm0.039$.
\citet{bonifacio_search_2000} found however that the
\citet{schlegel_maps_1998} maps overestimate the reddening values when
the color excess exceeds about 0.10\,mag. We therefore applied the
correction proposed by \citet{bonifacio_search_2000}, and additionally
corrected for the mean cluster distance by a factor
$\exp(-\abs{d\sin(b)}/h)$ where $d$ is the average cluster distance,
$b$ is the average galactic latitude, and $h$ is the scale height of
the galactic disk, assumed to be 125\,pc.  This yielded our adopted
$E(\mathrm{B}-\mathrm{V})=0.102\pm0.019$, where the uncertainty
reflects the standard deviation of the individual
\citet{schlegel_maps_1998} values.  Finally, to convert to Gaia
colors we used the calibration of \citet{stassun_TIC8_2019}, namely
$E(\mathrm{Bp}-\mathrm{Rp})=1.31 E(\mathrm{B}-\mathrm{V})$ and
$A_G=2.72 E(\mathrm{B}-\mathrm{V})$.  This yielded
$E(\mathrm{Bp}-\mathrm{Rp})=0.134\pm0.025$, which is used in the
plots.  To redden the isochrones, we assumed $R_V=3.1$, and applied
the \citet{odonnell_1994} SED-dependent extinction law star-by-star
through the PARSEC interface.

For the metallicity, we considered a range of super and sub-solar
metallicities to fit as much of the locus from
$0.5<\mathrm{Bp}-\mathrm{Rp}<1.5$ as possible, and settled on the
slightly super-solar $[M/H]=0.06$ \citep{cummings_2011_li_iron}.
Sub-solar metallicities led to model predictions too blue along the
main sequence by $\approx$0.1\,mag.  Our adopted metallicity agrees
with the spectroscopic metallicity from
\citet[][Sec~4.4.4]{cummings_2011_li_iron}, and with the iron
abundance recently determined by the Gaia-ESO team
\citep{baratella_gaiaeso_2020}, which represented an up-revision from
an earlier sub-solar metallicity \citep{randich_gaiaeso_2018}.  We
caution though that other sub-solar metallicities have been reported
\citep{bailey_rv_2018}, and note that the cluster metallicity and
reddening are degenerate; if the reddening is lowered, the implied
metallicity will increase.  While a detailed redetermination of
these parameters from the Gaia photometry is beyond our scope, our
adopted values for both the reddening and metallicity are within the
range of previously reported values in the literature.

@@

        \vspace{-0.5cm}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{f3c.pdf}
			\includegraphics[width=0.49\textwidth]{f3d.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf TESS rotations periods and dereddened Gaia colors for
  \cn.} {\it Top row}: \cn\ split into its core and halo.  Sets
  $\mathcal{A}$ ({\it left}) and $\mathcal{B}$ ({\it right}) undergo
  successive stages of automated cleaning, (see
  Section~\ref{subsubsec:cluster}).  The Pleiades
  \citep[125\,Myr;][]{rebull_rotation_2016a} and Praesepe
  \citep[650\,Myr;][]{douglas_poking_2017} are shown for reference.
  Stars in the core and the halo of NGC\,2516 overlap with the
  Pleiades at $\bpmrp$$<$1.2.  From $1.2<\bpmrp<1.7$, some NGC\,2516
  members have longer rotation periods than Pleaides members.  {\it
  Bottom row}: \cn\ colored by binarity indicators.  Sources with
  $\mathrm{RUWE}>1.2$ appear in red (astrometric binaries; 11\% of
  stars).  Overplotted in orange are sources at least 0.3\,mag above
  an empirically fit cluster isochrone (photometric binaries; 20\% of
  stars).
  Rotation periods for comparison field stars are shown in
  Appendix~\ref{app:compstar}.
  \label{fig:rot}
	}
\end{figure*}

@@

The resulting rotation periods and periodogram powers are reported,
regardless of detection significance, in Table~\ref{tab:maintable}.
To clean the resulting rotation period measurements, we designed a set
of automated cleaning criteria, which yielded sets we denote
$\mathcal{A}$ and $\mathcal{B}$.  For both sets, we considered light
curves for which the peak Lomb-Scargle periodogram period was below 15
days, the normalized Lomb-Scargle power exceeded 0.08, and for which
no equal-brightness or greater companions were within the aperture.
These limits were chosen after inspecting the light curves visually,
to eliminate low-quality rotation periods while retaining high
completeness.  Beyond requiring that the target star be the brightest
star in the aperture, we also required that at most one companion with
brightness exceeding one tenth of the target star's brightness be
present in the aperture according to the Gaia DR2 source catalog.  For
set $\mathcal{A}$, this yielded \nautorotnumerator\ light curves, out
of \nautorotdenominator\ stars for which rotation periods might
plausibly have been detected.  For set $\mathcal{B}$, we additionally
required that the Lomb-Scargle and Stellingwerf phase-dispersion
methods yielded identical periods to within 10\%.  This yielded
\nautorotnumeratormatching\ light curves.  Boolean flags for each set
are included in Table~\ref{tab:maintable}.

@@

The results are shown in the top two rows of
Figure~\ref{fig:physical_x_rotn}.  In these figures, to ensure a fair
comparison, we only show stars with $0.5<\bpmrp<1.2$ for which our
TESS pipeline succeeding in making light curves.  In other words, the
stars in the base sample are those for which we could have plausibly
detected a rotation period.  This color range is preferred because the
the slow sequence becomes less defined for spectral types later than
$\approx$K4V.  The bins in the histograms (top-right) were chosen to
ensure that the same number of stars were in each bin.  Rotation
periods are detected for 66\% of stars at least 25\,pc from the
cluster core.  The average detection rate inside of 25\,pc, 73\%, is
slightly higher.  With $\approx$3.5\,pc, the period detection rate of
61\% is relatively low, due to the combined effects of crowding and
saturation.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.9\textwidth]{f5a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown versus
  dereddened color for all kinematic \cn\ members with Gaia-ESO
  ($N=452$) or GALAH ($N=107$) spectra available.  The GALAH spectra
  comprise slightly over half of the halo stars, due to the
  non-targeted selection function of that survey.  Field stars (gray
  points) are Keck-HIRES measurements of stars with detected
  transiting planets from the Kepler mission
  \citep{berger_identifying_2018}.  Stars in the Pleiades (125\,Myr)
  with Li detections are from \citet{bouvier_pleiades_lirot_2018};
  Praesepe detections (650\,Myr) are shown in violet
  \citep{soderblom_praesepe_1993}.  Points with ${\rm
  EW}\approx0\,$m\AA\ are non-detections.
  The Li measurements for both Core and Halo stars in \cn\ are
  consistent with a near-Pleiades age.
  {\bf LGB TODO: the increasing EWs for the Gaia-ESO M dwarfs must be erroneous. What's happening there?}
  \label{fig:lithiumcorehalo}
  }
\end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f5b.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium and rotation in NGC\,2516.} Points show all
  kinematic \cn\ members with TESS rotation periods in the Gaia-ESO or
  GALAH surveys, and are colored by equivalent width.  Points with
  reported equivalent widths below 20\,m\AA\ (including
  non-detections) are shown with gray crosses.  The lithium-rotation
  correlation for the K-dwarfs has been observed in other young
  clusters, and is discussed in Section~\ref{subsec:lithium}.
	\label{fig:lithiumrot}
	}
\end{figure*}

@@

From their rotation period analysis of \cn,
\citet{fritzewski_rotation_2020} have recently demonstrated that
between 100 and 200\,Myr, young clusters (\cn, Blanco\,1, Psc\,Eri,
M\,35, M\,50, and the Pleaides) all show similar rotation period
distributions.  The FGK stars show a ``slow sequence'', a population
of ``rapid rotators'', and in some cases stars in the ``void'' between
the two populations ({\it e.g.,}, Psc\,Eri and \cn\ show more stars in
the void than the Pleiades).  The late M dwarfs with detected rotation
periods tend to rotate rapidly ($<3$\,days), though there may also be
late M dwarfs with slow ($>20$\,days) rotation periods.  Although
these late M dwarfs are beyond the magnitude limit of our analysis,
they have been explored in depth for \cn\ by both
\citet{Irwin_NGC2516_2007} and \citet{fritzewski_rotation_2020}.
While a detailed comparison of our results with theirs is beyond the
scope of this work, we emphasize that the main contribution of our
study beyond these previous analyses is not our sensitivity to new
ranges of spectral type or rotation period, but rather our sensitivity
to rotation periods across both the core and halo of \cn, and a
corresponding expansion in the number of stars for which we can derive
rotation periods.  In the following section, we therefore focus on
the origins of the features in the rotation-color diagram, and whether
our expanded rotation sample can provide any new insights.

@@

The other features of the rotation-color diagram are indications that
magnetic braking is not always the most important effect.  The
question of why the rapid rotators exist, for instance, has not yet
been definitively answered.  Models that introduce variable
core-envelope decoupling timescales can reproduce each type of
population \citep{Irwin_NGC2516_2007}, but the question then shifts to
why these decoupling timescales should differ for stars that are
otherwise identical.  Another possible explanation invokes a
spontaneous and random change of magnetic field topology that causes
the dynamo to couple more strongly with the stellar wind
\citep{brown_metastable_2014}.  Other processes could be external to the
star entirely: \citet{qureshi_signature_2018} for instance have
reported that mergers between giant planets and the star could explain
a non-neglible fraction of the rapid rotators.
